SIMD\slash SIMT-vectorized Sparse Chemical Kinetic Jacobian\slash Thermo-chemical Source Term Evaluation

### 2018-03-07 paper.tex
revise error discussion

As in our previous work~\cite{Niemeyer:2016aa}, the correctness of the Jacobian codes generated by \texttt{pyJac} was determined by comparison to the Jacobian obtained by automatic differentiation of the source rate subroutines generated by \texttt{pyJac} via the \texttt{Adept} software library~\cite{adept-v11,hogan2014fast}.
The reasons for this choice are explained fully in~\cite{Niemeyer:2016aa}, but broadly speaking automatic differentiation provides relatively fast, highly accurate Jacobian matrix evaluation with minimal additional programming effort.
The discrepancy between the analytical and automatic differentiated Jacobians for thermo-chemical state $k$, denoted by $\mathcal{J}_k$ and $\hat{\mathcal{J}}_k$ respectively, is determined by the relative error Frobenius norm over all Jacobian indicies $i, j$:
\begin{equation}
 \label{e:jac_error_base}
 E_{\text{rel}, k} = \left\lVert \frac{\hat{\mathcal{J}}_{ijk} - \mathcal{J}_{ijk}}{\hat{\mathcal{J}}_{ijk}} \right\rVert_{i,j,F}
\end{equation}

To avoid large relative errors of small nonzero Jacobian elements due to accumulation of floating point error, the Frobenius norm of the automatically differentiated Jacobian is calculated over all thermo-chemical states $k$:
\begin{equation}
 \label{e:thresh}
 \mathcal{T} = \left\lVert \mathcal{\hat{J}} \right\rVert_{k, F}
\end{equation}
The error statistics reported in this section are then based only on matrix elements where $\mathcal{J}_{ijk} \ge \frac{\mathcal{T}}{\mathcal{C}}$; this filtered form of~\cref{e:jac_error_base} is denoted $E_{\mathcal{C},k}$.
Finally, the Frobenius norm is calculated over all the states in the PaSR thermo-chemical condition database:
\begin{equation}
 \label{e:thresholded_error}
 E_{\mathcal{C}} = \left\lVert E_{\mathcal{C},k} \right\rVert_{k, F}
\end{equation}

It is noted that this error norm is quite different from the relative error Frobenius norm suggested by Anderson et al.~\cite{Anderson:1999aa} for quantifying the error of matrices in LAPACK:
\begin{align}
 \label{e:error_lapack}
 \lhs{lapack}{E_{\mathcal{L}, k}} &=  \rhs{lapack}{\frac{\left\lVert \hat{\mathcal{J}}_{ijk} - \mathcal{J}_{ijk} \right\rVert_{i,j,F}}{\left\lVert \hat{\mathcal{J}}_{ijk} \right\rVert_{i,j,F}}} \nonumber \\
 \lhs{lapack}{E_{\mathcal{L}}} &= \rhs{lapack}{\left\lVert  E_{\mathcal{L}, k} \right\rVert_{k, F}}[c]
\end{align}
In our experience, the LAPACK error norm is often dominated by the accuracy of the larger elements in the Jacobian, while our filtered error norm (while significantly more stringent) is useful for identifying errors in smaller elements.
For reference, both our own error norm and the LAPACK error norm will be reported.

\begin{table}[htbp]
\sisetup{retain-zero-exponent=true}
\centering
\begin{tabular}{@{}l S[table-format=.0] S[table-format=1.3e1] S[table-format=1.3e1] S[table-format=1.3e1] S[table-format=1.3e1] @{}}
\toprule
Model                 & \multicolumn{1}{c}{$E_{\mathcal{L}}$} & \multicolumn{1}{c}{$\bar{\mathcal{T}}$} & \multicolumn{1}{c}{$E_{10^{20}}$}   & \multicolumn{1}{c}{$E_{10^{15}}$} \\
\midrule
\ce{H2}\slash \ce{CO} & \num{1.048e-11}      & \num{6.431e+18}      & \num{1.741e+0}  & \num{4.508e-5} \\
GRI-Mech 3.0          & \num{1.124e-14}      & \num{7.783e+19}      & \num{3.842e-7}  & \num{3.687e-7} \\
USC-Mech II           & \num{9.632e-15}      & \num{2.830e+21}      & \num{1.119e-2}  & \num{1.983e-7} \\
\ce{iC5H11OH}         & \num{1.227e-10}      & \num{2.733e+26}      & \num{1.363e-3}  & \num{2.764e-5} \\
\bottomrule
\end{tabular}
\caption{Summary of Jacobian matrix validation results.
Error statistics are the largest filtered relative error $E_\mathcal{C}$ for over all samples.
It is noted that the threshold described in~\cref{e:thresh} varies slightly between the \conp/ and \conv/ cases; the reported $\bar{\mathcal{T}}$ is the average of the two, however the appropriate value was used during calculations of the error statistics.
}
\label{T:error}
\end{table}

\Cref{T:error} reports the maximum $E_{\mathcal{C}}$ and $E_{\mathcal{L}}$ values over all test platforms\slash vectorization patterns (see~\cref{t:platforms}) and the sparse and dense Jacobians (see~\cref{S:sparsity}).
The most stringent filtered error norm ($\mathcal{C} = 10^{20}$) ranges from \numrange[retain-zero-exponent=true]{e-7}{e0}; the largest error is for the \ce{H2}\slash\ce{CO} model.
For this model, $\mathcal{T}$  is smaller than the tolerance of \num{e20}, and hence the error norm considers smaller than $\mathcal{O}(1)$.
It is noted that for GRI-Mech 3.0, which has a $\mathcal{T}$ roughly an order of magnitude larger, the stringent error norm is significantly smaller: $\mathcal{O}(\num{e-7})$.
Given the intricacy of floating point error evaluation (see~\cref{s:validation}), the use of different languages and OpenCL platforms (the effect of these differences will be explored in Appendix \textbf{???}\todo{this}), and the general complexity of \texttt{pyJac} it would be exceedingly difficult to pinpoint an exact cause for this phenomena.
Indeed, the stringent filtered error norm was chosen specifically for its ability to detect errors in relatively smaller Jacobian entries.
To ensure there is no bug or error present in \texttt{pyJac}, a relaxed filtered error norm ($\mathcal{C} = 10^{15}$) is also presented for each model in~\cref{T:error}.
This relaxed norm is smaller by \numrange{2}{5} orders of magnitude for all models---except GRI-Mech 3.0, where the stringent case has very small error as previously discussed---indicating that accuracy is higher when the smaller Jacobian entries are excluded.

The relative LAPACK error norm---ranging from \textasciitilde\numrange{e-11}{e-14}---re-enforces this finding, as it indicates roughly \numrange{11}{14} digits of accuracy~\cite{Anderson:1999aa}.
Interestingly, the LAPACK error norm does not correlate well with the more stringent filtered error norm, e.g., USC-Mech II has the lowest LAPACK error norm (\num{9.632e-15}), but the second largest stringent filtered error norm (\num{1.119e-2}).
Conversely, the LAPACK error norm does seem to correlate with the relaxed filtered error norm; the two models with the largest LAPACK error, \ce{H2}\slash\ce{CO} and \ce{iC5H11OH}, have the largest relaxed filtered error norms ($\mathcal{O}(\num{e-5})$).
This study, along with the individual unit-testing of all chemical source terms and Jacobian sub-components described in~\cref{s:unittest}, gives us high confidence in the correctness of this new version of \texttt{pyJac}.

>>>>

As in our previous work~\cite{Niemeyer:2016aa}, the correctness of the Jacobian was determined by comparison to the Jacobian obtained by automatic differentiation of the source rate subroutines generated by \texttt{pyJac} via the \texttt{Adept} software library~\cite{adept-v11,hogan2014fast}.
The reasons for this choice are explained fully in~\cite{Niemeyer:2016aa}, but broadly speaking automatic differentiation provides relatively fast, highly accurate Jacobian matrix evaluation with minimal additional programming effort.
The discrepancy between the analytical and automatic differentiated Jacobians for thermo-chemical state $k$, denoted by $\mathcal{J}_k$ and $\hat{\mathcal{J}}_k$ respectively, is determined by the relative error Frobenius norm over all Jacobian indicies $i, j$:
\begin{equation}
 \label{e:jac_error_base}
 E_{\text{rel}, k} = \left\lVert \frac{\hat{\mathcal{J}}_{ijk} - \mathcal{J}_{ijk}}{\hat{\mathcal{J}}_{ijk}} \right\rVert_{i,j,F}
\end{equation}

To avoid large relative errors of very small nonzero Jacobian elements due to accumulation of floating point error, the Frobenius norm of the automatically differentiated Jacobian is calculated over all thermo-chemical states $k$:
\begin{equation}
 \label{e:thresh}
 \mathcal{T} = \left\lVert \mathcal{\hat{J}} \right\rVert_{k, F}
\end{equation}
The error statistics reported in this section are then based only on matrix elements where $\mathcal{J}_{ijk} \ge \frac{\mathcal{T}}{\mathcal{C}}$; this filtered form of~\cref{e:jac_error_base} is denoted $E_{\mathcal{C},k}$.
Finally, the Frobenius norm is calculated over all the states in the PaSR thermo-chemical condition database:
\begin{equation}
 \label{e:thresholded_error}
 E_{\mathcal{C}} = \left\lVert E_{\mathcal{C},k} \right\rVert_{k, F}
\end{equation}

It is noted that this error norm is quite different from the relative error Frobenius norm suggested by Anderson et al.~\cite{Anderson:1999aa} for quantifying the error of matrices in LAPACK:
\begin{align}
 \label{e:error_lapack}
 \lhs{lapack}{E_{\mathcal{L}, k}} &=  \rhs{lapack}{\frac{\left\lVert \hat{\mathcal{J}}_{ijk} - \mathcal{J}_{ijk} \right\rVert_{i,j,F}}{\left\lVert \hat{\mathcal{J}}_{ijk} \right\rVert_{i,j,F}}} \nonumber \\
 \lhs{lapack}{E_{\mathcal{L}}} &= \rhs{lapack}{\left\lVert  E_{\mathcal{L}, k} \right\rVert_{k, F}}[c]
\end{align}
In our experience, the LAPACK error norm is often dominated by the accuracy of the larger elements in the Jacobian, while the filtered error norm (although significantly more stringent) is useful for identifying errors in both large and small Jacobian entries.
Further, with a tunable threshold parameter, $\mathcal{C}$, the error of different ranges of of element sizes can be assessed, and the effects of floating point error can be isolated.
For reference, both our own error norm and the LAPACK error norm will be reported.

\begin{table}[htbp]
\sisetup{retain-zero-exponent=true}
\centering
\begin{tabular}{@{}l S[table-format=.0] S[table-format=1.3e1] S[table-format=1.3e1] S[table-format=1.3e1] S[table-format=1.3e1] @{}}
\toprule
Model                 & \multicolumn{1}{c}{$E_{\mathcal{L}}$} & \multicolumn{1}{c}{$\bar{\mathcal{T}}$} & \multicolumn{1}{c}{$E_{\mathcal{C} = 10^{20}}$}   & \multicolumn{1}{c}{$E_{\mathcal{C} = 10^{15}}$} \\
\midrule
\ce{H2}\slash \ce{CO} & \num{1.048e-11}      & \num{6.431e+18}      & \num{1.741e+0}  & \num{4.508e-5} \\
GRI-Mech 3.0          & \num{1.124e-14}      & \num{7.783e+19}      & \num{3.842e-7}  & \num{3.687e-7} \\
USC-Mech II           & \num{9.632e-15}      & \num{2.830e+21}      & \num{1.119e-2}  & \num{1.983e-7} \\
\ce{iC5H11OH}         & \num{1.227e-10}      & \num{2.733e+26}      & \num{1.363e-3}  & \num{2.764e-5} \\
\bottomrule
\end{tabular}
\caption{Summary of Jacobian matrix validation results.
Error statistics are the largest filtered relative error $E_\mathcal{C}$ for over all test platforms, vectorization patterns (\cref{t:platforms}),  \conp/\slash \conv/ and sparse\slash dense Jacobians.
It is noted that the threshold described in~\cref{e:thresh} varies slightly between the \conp/ and \conv/ cases; the reported $\bar{\mathcal{T}}$ is the average of the two, however the appropriate value was used during calculations of the error statistics.
}
\label{T:error}
\end{table}

\Cref{T:error} reports the maximum $E_{\mathcal{C}}$ and $E_{\mathcal{L}}$ values over all test platforms\slash vectorization patterns (see~\cref{t:platforms}), sparse\slash dense (see~\cref{S:sparsity}) and \conp/\slash\conv/ Jacobians.
The most stringent filtered error norm ($\mathcal{C} = 10^{20}$) ranges from \numrange[retain-zero-exponent=true]{e-7}{e0}; the largest error is for the \ce{H2}\slash\ce{CO} model.
For this model, $\mathcal{T}$  is smaller than the tolerance of \num{e20}, and hence the error norm considers smaller than $\mathcal{O}(1)$.
It is noted that for GRI-Mech 3.0, which has a $\mathcal{T}$ roughly an order of magnitude larger, the stringent error norm is significantly smaller: $\mathcal{O}(\num{e-7})$.
Given the intricacy of floating point error evaluation, the use of different languages and OpenCL platforms (the effect of these differences will be explored in Appendix \textbf{???}\todo{this}), and the general complexity of \texttt{pyJac} it would be exceedingly difficult to pinpoint an exact cause for this phenomena, as was done in~\cref{s:validation}.
To ensure there is no bug or error present in the Jacobians generated by \texttt{pyJac}, a relaxed filtered error norm ($\mathcal{C} = 10^{15}$) is also presented for each model in~\cref{T:error}.
This relaxed norm is smaller by \numrange{2}{5} orders of magnitude for all models---except GRI-Mech 3.0, where the stringent case has very small error as previously discussed---indicating that accuracy is higher when the smaller Jacobian entries are excluded; this suggests floating point error accumulation is the primary driver of the stringent filtered error norm.

The relative LAPACK error norm---ranging from \textasciitilde\numrange{e-10}{e-14}---re-enforces this finding, as it indicates roughly \numrange{10}{14} digits of accuracy~\cite{Anderson:1999aa}.
Further, the LAPACK error norm does not correlate well with the stringent filtered error norm, e.g., USC-Mech II has the lowest LAPACK error norm (\num{9.632e-15}), but the second largest stringent filtered error norm (\num{1.119e-2}).
Conversely, the LAPACK error norm does seem to correlate with the relaxed filtered error norm; the two models with the largest LAPACK error, \ce{H2}\slash\ce{CO} and \ce{iC5H11OH}, have the largest relaxed filtered error norms ($\mathcal{O}(\num{e-5})$), again suggesting that the stringent error norm is influenced by floating point error accumulation.
These findings, along with the individual unit-testing of all chemical source terms and Jacobian sub-components described in~\cref{s:unittest}, gives high confidence in the correctness of this new version of \texttt{pyJac}.
